---
title: "Economic Papers"
author: "Sergio Aquino"
date: 2022-09-06
output: html_output
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

# A Brief Look at James J. Heckman's papers in the National Bureau of Economic Research (NBER)

James J Heckman, born in 1944, is a Nobel Prize winning American economist at the University of Chicago. Heckman is noted for his contributions to selection bias and self-selection analysis, especially Heckman correction, which earned him the Nobel Prize in Economics. He is also well known for his empirical research in labor economics, particularly regarding the efficacy of early childhood education programs.

His recent research focuses on inequality, human development and lifecycle skill formation, with a special emphasis on the economics of early childhood education. He is currently conducting new social experiments on early childhood interventions and reanalyzing old experiments. He is also studying the emergence of the underclass in the US and Western Europe. For example, he showed that a high IQ only improved an individual's chances of financial success by 1 or 2%. Instead, "conscientiousness," or "diligence, perseverance and self-discipline," are what led to financial success.

The National Bureau of Economic Research (NBER) was founded in 1920, largely in response to heated Progressive-era controversies over income distribution. The two leading figures in its launch — Malcolm Rorty, an executive at the American Telephone and Telegraph Company, and Nachum Stone, a socialist labor organizer with a PhD in economics from Columbia University — had widely different views on many economic policy issues. They agreed, however, that there was little data on which to base discussions of these issues. With the support of a group of business and labor leaders, as well as university-based economists who were committed to uncovering and disseminating important facts about the economy of the United States, they created the NBER to address this information gap. 

To this day, NBER research is bound by a restriction that the founders imposed: studies may present data and research findings, but may not make policy recommendations or make normative statements about policy.  

# Load libraries

```{r , include=FALSE}
#install.packages("nberwp")

library(nberwp)
library(tidyverse)
library(ggthemes)
library(tidytext)
library(SnowballC)
library(RColorBrewer)
```


# Load and explore data

```{r}
authors <- nberwp::authors
papers <- nberwp::papers
paper_authors <- nberwp::paper_authors
```


## Join table and select papers that were authored by James J Heckman

Paper authors is a table that links papers to authors.

```{r}
j.heckman <- authors$author[str_detect(authors$name, "James J Heckman")]

heckman.papers <- paper_authors %>% 
  filter(author == j.heckman) %>% 
  inner_join(papers)  %>% 
  select(year, title) 
```

This is a very useful dataset for an economist, but for now, I'll simply plot the number of papers per year, and the most common words used in the titles.

It's interesting to note that Heckman didn't publish in 2001 and 2003 after having published often from 1998 to 2000. There is no hiatus from 2004 to 2021. 

## Plot number of papers per year

```{r}
heckman.papers %>%
  group_by(year) %>% 
  summarize(total = sum(n())) %>% 
  arrange() %>% 
  ggplot(aes(year, total)) +
    geom_col() +
    labs(
      x = "",
      y = "Number of Papers",
      title = "Number of Papers Published by James J. Heckman - 1974 to 2021",
      caption = "Sergio Aquino | Data: NBER"
  ) +
  theme_excel_new() + 
  scale_y_continuous(limits = c(0, 10), 
                     breaks = c(0,2,4,6,8,10))
  

```

## Most Common Base Words in the Titles of Heckman Papers

The base words are common in economics. Additionally, they show Heckman's research in labour economics with an emphasis in the efficacy of early childhood education programs.


```{r}
title.words <- heckman.papers %>%
  unnest_tokens(word, title) %>% 
  anti_join(stop_words, by = "word") %>% 
  mutate(stem = wordStem(word)) %>% 
  count(stem, sort = TRUE) %>%
  head(20) %>%
  mutate(stem = fct_reorder(stem, n)) %>% 
  ggplot(aes(stem,n)) +
    geom_col() +
    coord_flip() +
  labs(
      x = "Number of Base Words Used in Papers",
      y = "Stem",
      title = "Most Common Base Words in the Titles of Heckman Papers - 1974 to 2021",
      caption = "Sergio Aquino | Data: NBER"
  ) +
  theme_excel_new() +
  scale_y_continuous(limits = c(0, 30), 
                     breaks = c(0,10,20,30))

title.words
```



